---
title: "Group 6 CA2"
format: pdf
editor: visual
echo: false
warning: false
message: false
---

```{r}
library(tidyverse)
library(dplyr)
library(knitr)
library(tidyr)

X <- read.csv("CA2.csv")
X <- as.matrix(X)  

svd_X <- svd(X)

# Initialize lists to store approximations and errors
approximations <- list()
errors <- list()
frobenius_norms <- numeric(12)

# Compute lower-rank approximations and errors
for (k in 1:12) {
  U_k <- svd_X$u[, 1:k, drop = FALSE]  # Ensure correct matrix dimensions
  D_k <- diag(svd_X$d[1:k], nrow = k, ncol = k)  # Make it square
  V_k <- svd_X$v[, 1:k, drop = FALSE]
  
  X_k <- U_k %*% D_k %*% t(V_k)  # Compute rank-k approximation
  approximations[[k]] <- X_k
  errors[[k]] <- X - X_k
  frobenius_norms[k] <- sqrt(sum(errors[[k]]^2))
}
```

### Question 1:

```{r}
# 1. Mean vector of the rank 4 approximation error
delta_4_mean <- round(colMeans(errors[[4]]),2)
print("Mean Vector of Rank 4 Approximation Error:")
print(delta_4_mean)
```

### Question 2:

```{r}
# 2. Compare correlation matrices
cor_X <- round(cor(X),2)
cor_X2 <- round(cor(approximations[[2]]),2)
print("Correlation Matrix of X:")
print(cor_X)
print("Correlation Matrix of X_2:")
print(cor_X2)
```

###### Interpretation:

###### The rank-2 approximation exaggerates correlations, preserving major trends with values closer to -1 and 1. This suggests that the first two singular components capture the dominant variance, but finer variance details are lost (distorts smaller correlations), leading to over-simplified relationships.

### Question 3:

```{r}
# 3. Plot Frobenius norm as a function of k
plot(1:12, frobenius_norms, type="b", pch=19, col="blue", 
     xlab="Rank k", ylab="Frobenius Norm", 
     main="Frobenius Norm vs Rank")
```

###### Interpretation:

###### The Frobenius norm decreases as K increases, indicating that higher-rank approximations provide better approximations of X. The error reduction is most significant for lower values of K (e.g., from K=1 to K=6), suggesting that the first few singular components capture most of the data's variance. After K=8, the reduction in the Frobenius norm slows down. This means additional components contribute less to improving the approximation. The norm approaches zero at K=12, confirming that the full-rank approximation perfectly reconstructs X.

### Question 4:

```{r}
# 4. Plot percentage of total variation retained
variance_retained <- cumsum(svd_X$d^2) / sum(svd_X$d^2) * 100
plot(1:12, variance_retained[1:12], type="b", pch=19, col="red", 
     xlab="Rank k", ylab="Percentage Variance Retained", 
     main="Variance Retained vs Rank")
```

###### Interpretation:

###### The first few singular values capture a large proportion of the total variance. As K increases, the additional variance retained slows down and beyond K=8, additional components contribute less marginal improvement. At K=12, nearly all the variation is captured, meaning again that the full-rank matrix perfectly reconstructs X.
