---
title: "Group 17"
subtitle: "Continuous Assessment 1"
author: "LBNSTE002, AZRRAP001, RKHRIT001"
date: "2025-02-24"
format: pdf
editor: visual
execute: 
  echo: FALSE
  message: FALSE
  warning: FALSE
---

## Question 1:

```{r}

library(dplyr)

# Load the dataset
data <- read.csv("CA1.csv")

# Compute sample means for each time period
mean_vectors <- data |> 
  group_by(TimePeriod) |> 
  summarise(across(MaxBreadth:NasHeight, mean))

# Display the mean vectors
print(mean_vectors)

```

## Question 2:

```{r}

# Load necessary libraries
library(dplyr)
library(corrplot)

# Load the dataset correctly
CA1 <- read.csv("CA1.csv") # Ensure this matches the actual file name

# Period One
period_one <- CA1 |> filter(TimePeriod == 1)
R1 <- cor(period_one[,1:4])
corrplot(R1, method = "number")
mtext("Time Period 1 Correlation Matrix", side = 2, line = 1, cex = 1, las = 3)

# Period Two
period_two <- CA1 |> filter(TimePeriod == 2)
R2 <- cor(period_two[,1:4])
corrplot(R2, method = "number")
mtext("Time Period 2 Correlation Matrix", side = 2, line = 1, cex = 1, las = 3)

# Period Three
period_three <- CA1 |> filter(TimePeriod == 3)
R3 <- cor(period_three[,1:4])
corrplot(R3, method = "number")
mtext("Time Period 3 Correlation Matrix", side = 2, line = 1, cex = 1, las = 3)

# Period Four
period_four <- CA1 |> filter(TimePeriod == 4)
R4 <- cor(period_four[,1:4])
corrplot(R4, method = "number")
mtext("Time Period 4 Correlation Matrix", side = 2, line = 1, cex = 1, las = 3)

# Period Five
period_five <- CA1 |> filter(TimePeriod == 5)
R5 <- cor(period_five[,1:4])
corrplot(R5, method = "number")
mtext("Time Period 5 Correlation Matrix", side = 2, line = 1, cex = 1, las = 3)

# Adjust the layout to show multiple plots (optional)
par(mfrow = c(3,2))

```

###### Over time, the correlation between Maximal Breadth (X1) and Basialveolar Length (X3) weakened, indicating a shift in skull proportions. The relationship between Basibregmatic Height (X2) and Nasal Height (X4) strengthened in later periods, suggesting structural changes. Additionally, some previously positive correlations turned negative, particularly between Basibregmatic Height (X2) and Maximal Breadth (X1) in later periods, showing evolving skull shape patterns.

## Question 3:

```{r}
# Compute deviation vectors for X1 (MaxBreadth) and X3 (BasLength) in Period 1
period_1_data <- subset(data, TimePeriod == 1)[, c("MaxBreadth", "BasLength")]
deviation_X1 <- period_1_data$MaxBreadth - mean(period_1_data$MaxBreadth)
deviation_X3 <- period_1_data$BasLength - mean(period_1_data$BasLength)

dot_product <- sum(deviation_X1 * deviation_X3)
norm_X1 <- sqrt(sum(deviation_X1^2))
norm_X3 <- sqrt(sum(deviation_X3^2))
angle_radians <- acos(dot_product / (norm_X1 * norm_X3))
angle_degrees <- angle_radians * (180 / pi)
print(paste("Angle between deviation vectors:", angle_degrees, "degrees"))
print(paste("Angle between deviation vectors:", angle_radians, "radians"))
```

###### The angle between the deviation vectors for MaxBreadth (X1) and BasLength (X3) in Period 1 is approximately 89.14, which is close to 90. This is expected because the correlation coefficient between these two variables in Period 1, as seen in the correlation matrix, is close to zero. A near-zero correlation indicates that the two variables are largely uncorrelated, meaning their deviation vectors are almost orthogonal, leading to an angle close to 90.

## Question 4

```{r}
# Calculate Yi for all time periods Yi = 3 * NasHeight - MaxBreadth
data <- data %>%
  mutate(Y = 3 * NasHeight - MaxBreadth)

# CHECK this with linear comb.
(y_means <- data %>%
  group_by(TimePeriod) %>%
  summarise(Y_mean = mean(Y)))

# Calculate the covariance matrix for Y
y_cov_matrix <- cov(data %>% select(Y))

#use b as linear combination 
b <- c(-1, 0, 0, 3) 

b <- c(-1, 0, 0, 3)
mean_t <- t(mean_vectors[1:5,2:5])
y_i <- b%*%(mean_t)

#Y_i
y1<- b%*%t(period_one[,1:4])
y2<- b%*%t(period_two[,1:4])
y3<- b%*%t(period_three[,1:4])
y4<- b%*%t(period_four[,1:4])
y5<- b%*%t(period_five[,1:4])

#Y = Y_i
Y <- matrix(c(y1, y2, y3, y4, y5), nrow = 30)
cov(Y)
```
